---
title: "Assignment 1 - Incomplete Data Analysis"
author: "Zijun Lin"
output: html_document
---
1.
(a). (ii)  
(b). (ii)  
(c). (iii)  

2. Each variable contains 100*10% = 10 missing values.
The largest subsample contains 90 subjects. It means that in 10 subjects, all variables are missing, while other 90 cases have no missingness.
The smallest subsample contains no subjects. It means that each case has and only has one missing variable, then the number of subjects with missingness is 10*10 = 100. So 100 subjects would be discarded under a complete case analysis. 

3.
```{r, include = TRUE, message = FALSE}
#generate z1, z2, z3, y1, y2
set.seed(1)
z1 <-  rnorm(500, 0, 1)
set.seed(11)
z2 <-  rnorm(500, 0, 1)
set.seed(111)
z3 <-  rnorm(500, 0, 1)

y <- data.frame(y1 = 1 + z1, y2 = 5 + 2*z1 + z2)
```

(a). The machanism is MAR because the missingness depends on Y1 but not depends on Y2 itself.
```{r, include = TRUE, message = FALSE}
a = 2
b = 0

#storing the "missing" Y2 values
ind_MAR <-  which(-(a*(y$y1-1) + b*(y$y2-5) + z3) < 0)
y2_obs_MAR <- y$y2[ind_MAR]
y1_obs_MAR <- y$y1[ind_MAR]

#fit the regression model
fit_MAR <- lm(y2 ~ y1, y)
fit_obs_MAR <- lm(y2_obs_MAR ~ y1_obs_MAR)
fit_MAR; fit_obs_MAR

plot(density(y$y2), col = "blue", lwd = 2, xlab = "Y2", ylim = c(0, 0.25), main = "MAR")
lines(density(y2_obs_MAR), lwd = 2, col = "red")
legend(8.5, 0.20, legend = c("Complete data", "Observed data"), col = c("blue", "red"), lwd = c(2, 2))

```

The figure of the observed data is right-skewed. It means that y2 is more likely to miss when y1 is small. 

(b).
```{r, include = TRUE, message = FALSE}
#simulate the missingness
y_obs_MAR <- y
y_obs_MAR$y2[-ind_MAR] <- NA

predicted_sri <- predict(fit_obs_MAR)

#creating the completed y2 variable
set.seed(1)
y2_sri_MAR <- ifelse(is.na(y_obs_MAR$y2), predicted_sri, y_obs_MAR$y2) + rnorm(nrow(y_obs_MAR), 0, sigma(fit_MAR))

#create marginal distribution of y2
fit_sri_MAR <- lm(y2_sri_MAR ~ y$y1)
fit_sri_MAR

plot(density(y$y2), col = "blue", lwd = 2, xlab = "Y2", ylim = c(0, 0.25), main = "MAR")
lines(density(y2_sri_MAR), lwd = 2, col = "red")
legend(8.5, 0.20, legend = c("Complete data", "Completed data"), col = c("blue", "red"), lwd = c(2, 2))
```




(c). The machanism is MAR because the missingness depends on Y2 itself.
```{r, include = TRUE, message = FALSE}
a = 0
b = 2

#storing the "missing" Y2 values
ind_MNAR <-  which(-(a*(y$y1-1) + b*(y$y2-5) + z3) < 0)
y2_obs_MNAR <- y$y2[ind_MNAR]
y1_obs_MNAR <- y$y1[ind_MNAR]

#fit the regression model
fit_MNAR <- lm(y2 ~ y1, y)
fit_obs_MNAR <- lm(y2_obs_MNAR ~ y1_obs_MNAR)
fit_MNAR; fit_obs_MNAR

plot(density(y$y2), col = "blue", lwd = 2, xlab = "Y2", ylim = c(0, 0.35), main = "MNAR")
lines(density(y2_obs_MNAR), lwd = 2, col = "red")
legend(8.5, 0.30, legend = c("Complete data", "Observed data"), col = c("blue", "red"), lwd = c(2, 2))
```

The figure of the observed data is right-skewed. It means that y2 is more likely to miss when y1 is small. 


(d).
```{r, include = TRUE, message = FALSE}
#simulate the missingness
y_obs_MNAR <- y
y_obs_MNAR$y2[-ind_MNAR] <- NA

predicted_sri <- predict(fit_obs_MNAR)

#creating the completed y2 variable
set.seed(1)
y2_sri_MNAR <- ifelse(is.na(y_obs_MNAR$y2), predicted_sri, y_obs_MNAR$y2) + rnorm(nrow(y_obs_MNAR), 0, sigma(fit_MNAR))

#create marginal distribution of y2
fit_sri_MNAR <- lm(y2_sri_MNAR ~ y$y1)
fit_sri_MNAR

plot(density(y$y2), col = "blue", lwd = 2, xlab = "Y2", ylim = c(0, 0.30), main = "MNAR")
lines(density(y2_sri_MNAR), lwd = 2, col = "red")
legend(8.5, 0.20, legend = c("Complete data", "Completed data"), col = c("blue", "red"), lwd = c(2, 2))
```
The MNAR mechanism is harder to handle. 

4.
```{r, include = TRUE, message = FALSE}
load("databp.Rdata")
```
(a)
```{r, include = TRUE, message = FALSE}
#indices of subjects with recover time observed
ind <- which(is.na(databp$recovtime) == FALSE)

# mean value under complete case analysis
mcca <- mean(databp$recovtime, na.rm = TRUE)
mcca

# standard error under complete case analysis
sdcca <- sd(databp$recovtime, na.rm = TRUE)/sqrt(length(ind))
sdcca

# correlations between the recovery time and the dose
correcov_dose_cca <- cor(databp$recovtime[ind], databp$logdose[ind], method = "pearson")

# correlations between the recovery time and the blood pressure
correcov_bloodp_cca <- cor(databp$recovtime[ind], databp$bloodp[ind], method = "pearson")
correcov_dose_cca; correcov_bloodp_cca
```
The mean value of the recovery time is 19.27273. The associated standard error is 2.603013. The correlations between the recovery time and the dose is 0.2391256. The correlations between the recovery time and blood pressure is -0.01952862. 


(b).
```{r, include = TRUE, message = FALSE}
sbpmi <- ifelse(is.na(databp$recovtime) == TRUE, mean(databp$recovtime, na.rm = TRUE), databp$recovtime)

n <- length(sbpmi)
# mean of the completed variables
mmi <- mean(sbpmi)
mmi

# associated standard error of the completed variables
sdmi <- sd(sbpmi)/sqrt(n)
sdmi

# correlations between the recovery time and the dose
correcov_dose_mi <- cor(sbpmi, databp$logdose, method = "pearson")

# correlations between the recovery time and the blood pressure
correcov_bloodp_mi <- cor(sbpmi, databp$bloodp, method = "pearson")
correcov_dose_mi; correcov_bloodp_mi

```
The mean value of the recovery time is 19.27273. The associated standard error is 2.284135. The correlations between the recovery time and the dose is 0.2150612. The correlations between the recovery time and blood pressure is -0.01934126. 

(c).
```{r, include = TRUE, message = FALSE}
#fit the regression model
fitsbp <- lm(recovtime ~ logdose + bloodp, data = databp)

predri <- predict(fitsbp, newdata = databp)

sbpri <- ifelse(is.na(databp$recovtime) == TRUE, predri, databp$recovtime)

# mean of the completed variables
mri <- mean(sbpri)
mri

# associated standard error of the completed variables
sdri <- sd(sbpri)/sqrt(n)
sdri

# correlations between the recovery time and the dose
correcov_dose_ri <- cor(sbpri, databp$logdose, method = "pearson")

# correlations between the recovery time and the blood pressure
correcov_bloodp_ri <- cor(sbpri, databp$bloodp, method = "pearson")
correcov_dose_ri; correcov_bloodp_ri
```
The mean value of the recovery time is 19.44428. The associated standard error is 2.312845. The correlations between the recovery time and the dose is 0.2801835. The correlations between the recovery time and blood pressure is -0.0111364. 


(d).
```{r, include = TRUE, message = FALSE}
set.seed(1)
n <- length(databp$recovtime)
predsri <- predict(fitsbp, newdata = databp) + rnorm(n, 0, sigma(fitsbp))

sbpsri <- ifelse(is.na(databp$recovtime) == TRUE, predsri, databp$recovtime)

# mean of the completed variables
msri <- mean(sbpsri)
msri

# associated standard error of the completed variables
sdsri <- sd(sbpsri)/sqrt(n)
sdsri

# correlations between the recovery time and the dose
correcov_dose_sri <- cor(sbpsri, databp$logdose, method = "pearson")

# correlations between the recovery time and the blood pressure
correcov_bloodp_sri <- cor(sbpsri, databp$bloodp, method = "pearson")
correcov_dose_sri; correcov_bloodp_sri
```
The mean value of the recovery time is 20.4598. The associated standard error is 2.444571. The correlations between the recovery time and the dose is 0.2284537. The correlations between the recovery time and blood pressure is -0.01786944. 

(e).
```{r, include = TRUE, message = FALSE}
#hot deck possible donors
ind_donors <- which(is.na(databp$recovtime == FALSE))
ind_donors

for (j in predri[ind_donors]){
  a = c()
  for (i in predri[-ind_donors]){
    diff = (i - j)^2
    a = c(a, diff)
  }
  print(which.min(a))
}
```
```{r, include = TRUE, message = FALSE}
donor4 <- 5
donor10 <- 2
donor22 <- 15

rec_hotdeck <- c(databp$recovtime[1:3], predri[[donor4]], databp$recovtime[5:9], predri[[donor10]], databp$recovtime[11:21], predri[[donor22]], databp$recovtime[23:25])
rec_hotdeck
```


(f). The advantage of predictive mean matching is more accurate by using the predicted value. The potential problem is that the predicted value is not real data. 